Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-11-21 analysis/paper/paper.Rmd
added code chunk for PCA, MANOVA, and CV tests

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influnced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.

@@

# Isodeformation lines
tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
        amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
        palette = col_summer, grid = F)

>>>>

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.

@@

# Isodeformation lines
tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
        amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
        palette = col_summer, grid = F)
```

```{r PCA-analysis}
# pca
# these are fast
op.p <- PCA(op)

# We can see the pots groups by phase, and their shape distribution
# according to the first two principle components of the EFA 
# PC1 and PC2 explain 79% of variation
png("index/data/pottery-shape/pca-biplot.png", res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)

dev.off()

# different visualization 
plot(PCA(op), 2, pos.shp="full")
plot(PCA(op), 2, pos.shp="range")
plot(PCA(op), 2, pos.shp="xy")
plot(PCA(op), 2, pos.shp="range_axes")
plot(PCA(op), 2, pos.shp="full_axes")

# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  ggtitle("Shape variation along PC axes") +
  theme_minimal(base_size = 6)

ggsave("index/data/pottery-shape/pca-contrib.png")
scree_plot(op.p)

# PC1 is height of vessel
# PC2 is neck constriction
# PC3 is height of neck and roundness of body
```

```{r LDA-analysis}
# lda

# we performed a linear discriminant analysis (LDA) based on the new shape variables (PCs), with a leave-one-out cross-validation procedure, to identify the linear combination of shape features that was able to discriminate between phases

# we work on PCA scores
bot.l <- LDA(op.p, 2)
# print a summary, along with the leave-one-out cross-validation table. 
bot.l 
# plot.LDA works pretty much with the same grammar as plot.PCA
# here we only have one LD
plot(bot.l)
# plot the cross-validation table
plot_CV(bot.l)

# Does not produce an efficient separation between groups and a cross-validation (leave-one-out) at phase level showing a low confidence in the reclassification. But best for Pre-e, indicating these have the most distinctive shapes

#--------------------------------------------------
# Unsupervised model-based clustering

# what are the best grouping of the pots by shape?
library(mclust)

BIC <- mclustBIC(op.p$x[ , 1:5])
mod1 <- Mclust(op.p$x[ , 1:5], x = BIC)
summary(mod1, parameters = TRUE)
plot(mod1, what = "classification")

# no clusters apparent 
```

```{r MANOVA}
# MANOVA: Multivariate Analysis of (co)variace
# We can test for a difference in the distribution of PC scores by calculating a pairwise combination between every levels of a fac.

manv <-   MANOVA_PW(op.p, "phase")
manv$summary

#----------------------------------------------------
# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9

# Generalised additive mixed models (GAMMs) were used to explain shape variance with respect to mean environmental parameters and shell size, and to compare between individual shape features (PCs). A GAMM is a generalised linear mixed model with a linear predictor involving a sum of smooth functions of covariates. The model allowed for flexible specification of the dependence of the response on the covariates by defining regression splines (smoothing functions) and estimating their optimal degree of smoothness, rather than calculating parametric relationships

ggplot(pottery_data,
       aes(phase,
           as.numeric(Thick1_Neck))) +
  geom_boxplot()

pottery_data$thick_neck1 <- as.numeric(pottery_data$Thick1_Neck)
pottery_data$thick_body1 <- as.numeric(pottery_data$Thick1_Body)
pottery_data$height <- as.numeric(pottery_data$Hight)

plot(op.p$x[ , 2], pottery_data$thick_body1)

gams_input <- data.frame(thick_body1 = pottery_data$thick_body1 , 
                         height = pottery_data$height, 
                         shape_pc1 = op.p$x[ , 1])

library(brms)
m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
          data = gams_input, family = gaussian(), cores = 4, seed = 17,
          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
          control = list(adapt_delta = 0.99))
summary(m2) # so many zeros... not very promising 
# plot marginal effects for each predictor
plot(marginal_effects(m2), ask = FALSE)
loo(m2)
pp_check(m2)
```

```{r CV-test}
# CVs on shape PCs for standardisation

len <- length(op.p$x[,1])
# plot PC1-3 distribution by phase
pc_plot <- data_frame(pc = c(op.p$x[,1],
                             op.p$x[,2],
                             op.p$x[,3]),
                      phase = rep(op.p$fac$phase, 3),
                        pcn = str_glue('PC{sort(rep(1:3, len))}'))

pc_plot$phase <- fct_relevel(pc_plot$phase, 
                            c("Pre-European", 
                              "Post-European"))

ggplot(pc_plot,
       aes(phase,
           pc)) +
  geom_violin() +
  ggforce::geom_sina() +
  xlab("") +
  facet_wrap(~ pcn, ncol = 1) +
  theme_minimal()
ggsave("index/data/pottery-shape/pcs-boxplot-for-shape-variation.png",
       h = 10, w = 5)

# what are the variences for each PC in each period?
# we don't use CV because that is not good for measurements with no
# absolute zero 
pc_plot %>% 
  group_by(pcn, phase) %>% 
  summarise(variance = var(pc))

# statistical test
pc_plot_tests <- 
pc_plot %>% 
  nest_legacy(pc, phase) %>% 
  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$pc, .x$phase)))) %>% 
  unnest_legacy(var_test)

pc_plot_tests <- 
  pc_plot %>% 
  group_by_at(vars(-pc)) %>% 
  mutate(i = row_number()) %>% 
  ungroup() %>% 
  mutate(phase = as.character(phase),
         pcn = as.character(pcn)) %>% 
  spread(phase, pc) %>% 
  select(-i) %>% 
  nest_legacy(-pcn) %>% 
  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$`Post-European` ,
                                            .x$`Pre-European`)))) %>% 
  unnest(var_test)

library(cvequality)
pc1 <- data.frame(
x  = op.p$x[,1],
y = as.character(op.p$fac$phase),
stringsAsFactors = F)

cv_test_pc1 <- mslr_test(nr = 1000, pc1$x, pc1$y)

pc2 <- data.frame(
  x  = op.p$x[,2],
  y = as.character(op.p$fac$phase),
  stringsAsFactors = F)

cv_test_pc2 <- mslr_test(nr = 1000, pc2$x, pc2$y)

pc3 <- data.frame(
  x  = op.p$x[,3],
  y = as.character(op.p$fac$phase),
  stringsAsFactors = F)

cv_test_pc3 <- mslr_test(nr = 1000, pc3$x, pc3$y)

# do this pairwise
cv_test_pc1.1 <- 
  mslr_test(nr = 1000, 
            pc1$x[pc1$y %in% c("Post-European", 
                               "Pre-European")], 
            pc1$y[pc1$y %in% c("Post-European", 
                               "Pre-European")])

cv_test_pc1.2 <- 
  mslr_test(nr = 1000, 
            pc1$x[pc1$y %in% c("Chinese Contact", 
                               "Pre-European")], 
            pc1$y[pc1$y %in% c("Chinese Contact", 
                               "Pre-European")])

cv_test_pc1.3 <- 
  mslr_test(nr = 1000, 
            pc1$x[pc1$y %in% c("Chinese Contact", 
                               "Post-European")], 
            pc1$y[pc1$y %in% c("Chinese Contact", 
                               "Post-European")])

PC1_cv_tests <- 
bind_rows(cv_test_pc1.1, cv_test_pc1.2, cv_test_pc1.3) %>% 
  bind_cols(phases = c("Post-European vs Pre-European",
              "Chinese Contact vs Pre-European",
              "Chinese Contact vs Post-European"))

cv_test_pc2.1 <- 
  mslr_test(nr = 1000, 
            pc2$x[pc2$y %in% c("Post-European", 
                               "Pre-European")], 
            pc2$y[pc2$y %in% c("Post-European", 
                               "Pre-European")])

cv_test_pc2.2 <- 
  mslr_test(nr = 1000, 
            pc2$x[pc2$y %in% c("Chinese Contact", 
                               "Pre-European")], 
            pc2$y[pc2$y %in% c("Chinese Contact", 
                               "Pre-European")])

cv_test_pc2.3 <- 
  mslr_test(nr = 1000, 
            pc2$x[pc2$y %in% c("Chinese Contact", 
                               "Post-European")], 
            pc2$y[pc2$y %in% c("Chinese Contact", 
                               "Post-European")])

PC2_cv_tests <- 
  bind_rows(cv_test_pc2.1, 
            cv_test_pc2.2, 
            cv_test_pc2.3) %>% 
  bind_cols(phases = c("Post-European vs Pre-European",
                       "Chinese Contact vs Pre-European",
                       "Chinese Contact vs Post-European"))

bind_rows(PC1_cv_tests, 
          PC2_cv_tests, .id = "PC") %>% 
  mutate(PC = str_glue('PC{PC}')) %>% 
  mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
  ggplot(aes(str_wrap(phases, 40),
             p_value,
             colour = I(colour),
             shape = PC)
         ) +
  geom_point(size = 8,  position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 12) +
  coord_cartesian(clip = 'off') +
  xlab("") +
  ylab("CV equality test p-values")

ggsave("index/data/pottery-shape/cv-test-on-pcs.png", h = 3, w = 12)
